---
title: "Vignette Title"
author:
- name: Shouguo Gao
  affiliation: NHLBI/NIH
- name: Neal S. Young
  affiliation: NHLBI/NIH
  email: shouguo.gao@nih.gov
output:
  BiocStyle::html_document
abstract: |
  Description of your vignette
vignette: |
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette shows an example of applying pathway level to infer cell-type-specific co-expression networks and extracting co-expressed gene modules that are enriched for biological functions in cell types.

# 1. Load packages and data

```{r setup, warning = FALSE, message = FALSE}
library(CSCORE)
library(Seurat)
library(ggraph)
library(gridExtra)
source("../R/functions_2.R")
```

In this vignette, we use the single cell RNA-sequencing data on Peripheral blood mononuclear cells (PBMC) from COVID patients and healthy controls from [Wilk et al.](https://www.nature.com/articles/s41591-020-0944-y), which were also studied in our [manuscript](https://www.biorxiv.org/content/10.1101/2022.12.13.520181v1). This data set can be downloaded via the following bash script

```{r eval = TRUE}
#wget https://hosted-matrices-prod.s3-us-west-2.amazonaws.com/Single_cell_atlas_of_peripheral_immune_response_to_SARS_CoV_2_infection-25/blish_covid.seu.rds
cat("dowload")
```

After downloading blish_covid.seu.rds, we load it into the R session

```{r eval = TRUE}
####Load the gene set and convert into list
geneset<-geneIds(getGmt("../data/h.all.v2023.2.Hs.symbols.gmt"))
pbmc <- readRDS('../data/blish_covid.seu.rds')
# Use the original UMI counts stored in Assay 'RNA'
DefaultAssay(pbmc) <- 'RNA'
```

# 2. Select cell types and gene sets to study

In this example, we focus on CD4T cells and infer the CD4T cell-specific co-expression network.

```{r eval = TRUE}
pbmc_CD4T = pbmc[,pbmc$cell.type.coarse %in% 'CD4 T']
```

Depending on the biological question of interest, one may choose to study the co-expression network for any gene set. Here, we chose to infer the co-expression network for the genes with meaningful expression levels in CD4T cells (top 5000 among 26361 genes). There are several reasons for our choice:

1.  All genes with moderate to high expression levels provides a comprehensive and unbiased set of genes that could have meaningful biological functions in a cell type.

2.  If the genes have much lower expression levels, it would be statistically more challenging and biologically less interesting to infer their co-expressions, as these genes might have almost all UMI counts equal to 0.

In general, it will be up to the users's choice to select the gene sets to study. We recommend choosing the gene sets that are of interest to your application.

```{r eval = TRUE}
mean_exp = rowMeans(pbmc_CD4T@assays$RNA@counts/pbmc_CD4T$nCount_RNA)
genes_selected = names(sort.int(mean_exp, decreasing = T))[1:5000]
```

# 3. Run `Pathway Level` to infer cell-type-specific co-expression network on the specified gene set

We further subset the CD4 T cells to those from healthy control subjects in order to study CD4T-cell specific co-expression network among healthy control CD4T cells.

```{r eval = TRUE}
metaData<-pbmc_CD4T@meta.data
metaData_Healthy<-metaData[metaData$Status=="Healthy",]
metaData_COVID<-metaData[metaData$Status=="COVID",]
set.seed(666);metaData_COVID<-metaData_COVID[sample(1:dim(metaData_COVID)[1],dim(metaData_Healthy)[1]),]
```

# 4. Get the Subset of two groups

```{r eval = TRUE}
pbmc_CD4T.sub<-subset(pbmc_CD4T, cells=c(rownames(metaData_Healthy), rownames(metaData_COVID)))
rm(pbmc_CD4T)
```

# 5. Calculate the network, or score of networks


```{r eval = TRUE}
cor_data_score<-calculate_seq_correlation(geneset,pbmc_CD4T.sub,genes_selected,"COVID","Healthy", returnType = "score", scoreMethod = "GSCNCA")
save(cor_data_score, file = "CD4T_COVIDHD_cor_data_score.RData")
```

Next, based on the thresholded co-expression matrix, we apply [WGCNA](https://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-9-559) to extract co-expressed gene modules. In particular, we use `CS-CORE` estimates to measure co-expressions for single cell RNA-sequencing data, which replace the Pearson correlations used in traditional WGNCA workflow, that suffer from inflated false positives and attenuation bias on single cell data as demonstrated in our [manuscript](https://www.biorxiv.org/content/10.1101/2022.12.13.520181v1).

# 6. Calculate the network, or score of networks for permutations

```{r eval = TRUE}
cor_data_perm_list<-runPermute_seq_correlation(geneset,pbmc_CD4T.sub,genes_selected,"COVID","Healthy", ncores = 1, nPermute=20, returnType = "score", scoreMethod = "GSCNCA")
save(cor_data_perm_list, file = "CD4T_COVIDHD_cor_data_hallmark_perm_list_20.RData")
```

# 7. Calculate p values


```{r eval = TRUE}
pvalues<-sort(getPvaluesTTest(cor_data_score, cor_data_perm_list))
save(pvalues, file = "CD4T_pvaluesTtest.RData")
```

# 8. Generate the figures for top pathways

```{r eval = TRUE}
gglist <-list()
for(ii in 1:1){
  pathway<-names(pvalues)[ii]  
  networkslist<-network_seq_correlation(geneset[[pathway]],pbmc_CD4T.sub,genes_selected,"COVID","Healthy")
  gglist[[ii]]<-network_plot(networkslist,name1="COVID",name2="Healthy", fileName=paste0("CD4T_rank_",ii, "_", pathway, "_COVID_Healthy_Diff"))
}
```
```{r eval = TRUE}
for(ii in 2:2){
  pathway<-names(pvalues)[ii]  
  networkslist<-network_seq_correlation(geneset[[pathway]],pbmc_CD4T.sub,genes_selected,"COVID","Healthy")
  gglist[[ii]]<-network_plot(networkslist,name1="COVID",name2="Healthy", fileName=paste0("CD4T_rank_",ii, "_", pathway, "_COVID_Healthy_Diff"))
}
```
```{r eval = TRUE}
for(ii in 3:3){
  pathway<-names(pvalues)[ii]  
  networkslist<-network_seq_correlation(geneset[[pathway]],pbmc_CD4T.sub,genes_selected,"COVID","Healthy")
  gglist[[ii]]<-network_plot(networkslist,name1="COVID",name2="Healthy", fileName=paste0("CD4T_rank_",ii, "_", pathway, "_COVID_Healthy_Diff"))
}
```
```{r eval = TRUE}
for(ii in 4:4){
  pathway<-names(pvalues)[ii]  
  networkslist<-network_seq_correlation(geneset[[pathway]],pbmc_CD4T.sub,genes_selected,"COVID","Healthy")
  gglist[[ii]]<-network_plot(networkslist,name1="COVID",name2="Healthy", fileName=paste0("CD4T_rank_",ii, "_", pathway, "_COVID_Healthy_Diff"))
}
```

# 9. Show plots

Plot the first pathway.

```{r, fig.width = 20, fig.height = 8}
  gridExtra::grid.arrange(gglist[[1]]$gg1, gglist[[1]]$gg2, gglist[[1]]$ggDiff, ncol=3)
```

# 10. Show plots

Plot the second pathway.

```{r, fig.width = 20, fig.height = 8}
  library(ggraph)
  gridExtra::grid.arrange(gglist[[2]]$gg1, gglist[[2]]$gg2, gglist[[2]]$ggDiff, ncol=3)
```

# 11. Show plots

Plot the third pathway.

```{r, fig.width = 20, fig.height = 8}
  library(ggraph)
  gridExtra::grid.arrange(gglist[[3]]$gg1, gglist[[3]]$gg2, gglist[[3]]$ggDiff, ncol=3)
```

# 12. Show plots

Plot the fourth pathway.

```{r, fig.width = 20, fig.height = 8}
  library(ggraph)
  gridExtra::grid.arrange(gglist[[4]]$gg1, gglist[[4]]$gg2, gglist[[4]]$ggDiff, ncol=3)
```

The other Analysis is possible
# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
